---
title: "2D grids"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{grids2d}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

This vignette shows how to perform a density analysis on a plane.

This is only one of the possible analyzes that can be carried out by dividing the plane into a grid.

```{r setup}
library(rgrids)
library(ggplot2)
```

Doing a density analysis can be annoying in some circumstances. To assign points to each element of the grid in the worst case scenario, a double for loop is performed on rows and columns. Let's see how to make the process painless.

### Define dataframe of points
First, a set of random points is generated

```{r, fig.width = 6, fig.height = 4}
set.seed(1)

df_points <- data.frame(x = c(rnorm(n = 50000, mean = -2), 
                              rnorm(n = 50000, mean = 2)), 
                        y = c(rnorm(n = 50000, mean = 1), 
                              rnorm(n = 50000, mean = -1))
) 

head(df_points)

ggplot(df_points) + 
  geom_point(aes(x, y), color = "steelblue4", size = 0.1)
```

### Make 2D grid
Then a two-dimensional grid is built. A two-dimensional grid is defined by lower and upper bound along $x$ and $y$, and the number of cell along $x$ and $y$.

```{r}
# check the extreme values of the points along x
min(df_points$x)
max(df_points$x)

# check the extreme values of the points along y
min(df_points$y)
max(df_points$y)

# define boundaries of grid
(xmin <- floor(min(df_points$x)))
(xmax <- ceiling(max(df_points$x)))
(ymin <- floor(min(df_points$y)))
(ymax <- ceiling(max(df_points$y)))

# define the grid
grid2d <- makeGrid2d(
            xmin = xmin, xmax = xmax, xcell = 50,
            ymin = ymin, ymax = ymax, ycell = 50
          )
grid2d
```

In addition to the extremes and the number of cells, `makeGrid2d()` takes as input an additional parameter `by` which numbers the elements of the grid by increasing the $x$ (`h`) or the $y$ (`v`) faster:

```
by = "h"      by = "v"
7 8 9         3 6 9
4 5 6         2 5 8
1 2 3         1 4 7
```

In both cases the count starts from bottom left.

### Assign points to cells
Each point of the dataframe is assigned to its respective grid cell via `getCell()`.
The `getCell()` function takes as input an object of `Grid2d` class and a matrix or dataframe of points; if the passed matrix or dataframe has more than two columns, the first two will be automatically selected.

```{r}
grid_index <- getCell(grid2d, df_points)
df_points$grid_index <- grid_index

head(df_points)
```

The values of `grid_index` column range from 1 to 2500, which is the number of elements in the grid.


### Count the points
To get the occurrence of points in each cell just manipulate the previous result and get the grid coordinates. To facilitate the operation, use `getCounts()` function that takes `grid2d` and `grid_index` as input and returns a dataframe of three columns: the first two represent the coordinates of each element of the grid and the third represents the occurrence of points in each cell

```{r}
df_grid <- getCounts(grid2d, grid_index)

head(df_grid)
```

Finally, the grid is represented

```{r, fig.width = 6, fig.height = 4}
ggplot(df_grid) + 
  geom_raster(aes(x, y, fill = counts/nrow(df_grid))) + 
  theme(legend.title = element_blank())
```
