---
title: "1D grids"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{grids}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

This vignette shows how to obtain a boxplot or histogram given a dataframe of points in a plane.

```{r setup}
library(rgrids)
library(ggplot2)
```

## Boxplot

A density analysis in one dimension can be performed on points lying on a line suitably divided into segments, that is, different $x$ and $y = 0$, or on a set of points in the plane which is divided into strips.

If we consider a plane divided into strips, the grouping of points can be shown in a boxplot.  
Let's see how to proceed.

### Define dataframe of points
First we define a dataframe of random points

```{r, fig.width = 6, fig.height = 4}
set.seed(1)
    
df_points <- data.frame(x = c(rnorm(n = 5000, mean = -4), 
                              rnorm(n = 5000, mean = 4)), 
                        y = c(rnorm(n = 5000, mean = 1), 
                              rnorm(n = 5000, mean = -1))
) 
    
ggplot(df_points) + 
  geom_point(aes(x, y), color = "steelblue4", size = 0.1)
```

### Make 1D grid
Then a one-dimensional grid is built. A one-dimensional grid is defined by the lower bound, the upper bound, and the number of vertical stripes.

```{r}
# check the extreme values of the points along x
min(df_points$x)
max(df_points$x)
 
# define boundaries of grid
(xmin <- floor(min(df_points$x)))
(xmax <- ceiling(max(df_points$x)))
   
# define the grid
grid1d <- makeGrid1d(xmin = xmin, xmax = xmax, xcell = 16)
grid1d
```

The criterion for defining the grid boundaries and the number of strips varies from time to time based on the data and the type of analysis.

### Get the boxplot
Finally we get the boxplot using `getBoxplot()` function. The `getBoxplot()` function takes as input an object of class `Grid1d` and a dataframe of points. If the input dataframe has more than two columns, the first two will be automatically selected.
   
```{r}
df_boxplot <- getBoxplot(grid1d, df_points)
    
head(df_boxplot)
```

`getBoxplot()` assigns the same `x` value to all points within the same strip. The `xbp` column is a `factor`.

```{r, fig.width = 6, fig.height = 4}
ggplot(df_boxplot)+
  geom_boxplot(aes(x = xbp, y = y), fill = "steelblue3", color = "grey40") + 
  labs(x = "x") + 
  scale_x_discrete(breaks = levels(df_boxplot$xbp), drop = FALSE)
```

<!-- It is essential to add the `scale_x_discrete` line as in the example if you want to represent boxes without points (therefore empty spaces). By removing the line all the boxes are attached and the `x` axis shows jumps. -->

## Histogram
There are functions that can generate a histogram and provide all the information about it. However, a histogram can also be obtained using the `getCell()` and `getCounts()` function of `rgrids`.

`getCell()` function assigns each point of the plane to its respective strip, counting strips from left to right, while `getCounts()` function counts how many points fall within each strip and returns a dataframe with two columns: the center of the bin (i.e. the center of the strip) and the count.

```{r}
df_points$grid_index <- getCell(grid1d, df_points)
head(df_points)

df_hist <- getCounts(grid1d, df_points$grid_index)
head(df_hist)
```

```{r, fig.width = 6, fig.height = 4}
ggplot(df_hist) + 
  geom_col(aes(x, counts/sum(counts)), fill = "steelblue4") + 
  labs(y = "frequency")
```
